Loop-Aware Optimizations in PyPy's Tracing JIT

### 2011-07-01 talk/iwtc11/paper.tex
finetuning

\begin{abstract}
One of the nice properties of a tracing JIT is that many of its optimization
are simple requiring one forward pass. This is not true for loop-invariant code
motion which is a very important optimization for code with tight kernels.
In this paper we present a scheme for making simple optimizations loop-aware by
using a simple pre-processing step on the trace and not changing the
optimizations themselves. The scheme can give performance improvements of a
factor over two for PyPy's Python JIT executing simple numerical kernels
bringing the performance close to that of compiled C code.
\end{abstract}

@@

The iteration of the loop that was peeled off (lines 1-4) is called the
\emph{preamble}, the loop afterwards the \emph{peeled loop}.

@@

Before the trace is passed to a backend compiling it into machine code
it is optimized to achieve better performance.
One goal of that is to move 
operations out of the loop making them executed only once
and not every iteration. This we propose to achieve by loop peeling. It
leaves the loop body intact, but prefixes it with one iteration of the
loop. This operation by itself will not achieve anything. But if it is
combined with other optimizations it can increase the effectiveness of
those optimizations. For many optimization of interest some care has
to be taken when they are combined with loop peeling. This is
described below by explaining the loop peeling optimization
followed by a set of other optimizations and how they interact with
loop peeling.

@@

%XXX find reference of prior work on this

>>>>

\begin{abstract}
One of the nice properties of a tracing JIT is that many of its optimization
are simple requiring one forward pass only. This is not true for loop-invariant code
motion which is a very important optimization for code with tight kernels.
In this paper we present a scheme for making simple optimizations loop-aware by
using a simple pre-processing step on the trace and not changing the
optimizations themselves. The scheme can give performance improvements of a
factor over two for PyPy's Python JIT executing simple numerical kernels
bringing the performance close to that of compiled C code.
\end{abstract}

@@

The iteration of the loop that was peeled off (lines 1-4) is called the
\emph{preamble}, the loop afterwards (lines 6-9) the \emph{peeled loop}.

@@

Before a trace is passed to the backend compiling it into machine code
it is optimized to achieve better performance.
One goal of that is to move 
operations out of the loop making them executed only once
and not every iteration. This we propose to achieve by loop peeling. It
leaves the loop body intact, but prefixes it with one iteration of the
loop. This operation by itself will not achieve anything. But if it is
combined with other optimizations it can increase the effectiveness of
those optimizations. For many optimization of interest only a few
additional details has to be considered when they are combined with loop peeling. These are
described below by explaining the loop peeling optimization
followed by a set of other optimizations and how they interact with
loop peeling.

@@
